---
title: "Collaboration Networks in the Music Industry"
subtitle: "A Nested ERGM Analysis"
format: html
editor: visual
---

## Overview

This analysis examines the network of collaborations among the top 150 artists (by collaboration count) in the dataset. A series of nested Exponential Random Graph Models (ERGMs) is used to test four hypotheses:

1.  **Country homophily** — artists from the same country are more likely to collaborate.
2.  **Genre homophily** — artists sharing a similar musical genre are more likely to collaborate.
3.  **Collaboration activity** — highly collaborative artists are more likely to form additional ties.
4.  **Triadic closure** — artists who share collaborators are more likely to collaborate with each other.

## Setup

```{r}
#| label: setup
#| message: false

library(statnet)
library(dplyr)
library(network)
library(ergm)
```

## Data Loading & Preparation

```{r}
#| label: load-data

collab  <- read.csv("data/collaborations.csv")
artists <- read.csv("data/artists.csv")
```

We restrict the analysis to the top 150 artists by collaboration count, and subset the edge list to collaborations between artists in that group.

```{r}
#| label: subset-top-artists

top_artists <- artists %>%
  arrange(desc(collab_songs)) %>%
  slice(1:150)

collab_subset <- collab[
  collab$artist1 %in% top_artists$artist &
  collab$artist2 %in% top_artists$artist, ]
```

### Genre Grouping

Raw genre strings are collapsed into six broad genre groups.

```{r}
#| label: genre-groups

artists$genre_group <- "Other"

artists$genre_group[grepl("hip hop|rap|trap", artists$genres, ignore.case = TRUE)] <- "Hip-Hop"
artists$genre_group[grepl("latin|reggaeton", artists$genres, ignore.case = TRUE)] <- "Latin"
artists$genre_group[grepl("pop", artists$genres, ignore.case = TRUE)] <- "Pop"
artists$genre_group[grepl("electronic|dance|edm", artists$genres, ignore.case = TRUE)] <- "Electronic"
artists$genre_group[grepl("country", artists$genres, ignore.case = TRUE)] <- "Country"

table(artists$genre_group)
```

## Building the Network

```{r}
#| label: build-network

net <- network(collab_subset[, c("artist1", "artist2")], directed = FALSE)

# Country attribute
country_map <- setNames(artists$country, artists$artist)
net %v% "country" <- country_map[network.vertex.names(net)]

# Genre group attribute
genre_map <- setNames(artists$genre_group, artists$artist)
net %v% "genre_group" <- genre_map[network.vertex.names(net)]

# Collaboration count attribute
collab_map <- setNames(artists$collab_songs, artists$artist)
net %v% "collab_songs" <- collab_map[network.vertex.names(net)]
```

## Nested Model Building

Models are built incrementally, adding one term at a time, and compared using AIC/BIC to confirm each addition improves fit.

```{r}
#| label: model1

model1 <- ergm(net ~ edges)
```

```{r}
#| label: model2

model2 <- ergm(net ~ edges +
                  nodematch("country"))
```

```{r}
#| label: model3

model3 <- ergm(net ~ edges +
                  nodematch("country") +
                  nodematch("genre_group"))
```

```{r}
#| label: compare-1-3

AIC(model1, model2, model3)
BIC(model1, model2, model3)
```

```{r}
#| label: model4

model4 <- ergm(net ~ edges +
                  nodematch("country") +
                  nodematch("genre_group") +
                  gwesp(0.25, fixed = TRUE))
```

```{r}
#| label: compare-1-4

AIC(model1, model2, model3, model4)
BIC(model1, model2, model3, model4)
```

## Final Model

The final specification adds `nodecov("collab_songs")` to capture each artist's overall collaboration activity, alongside country and genre homophily and triadic closure.

```{r}
#| label: final-model

model5 <- ergm(net ~ edges +
                  nodematch("country") +
                  nodematch("genre_group") +
                  nodecov("collab_songs") +
                  gwesp(0.25, fixed = TRUE))

summary(model5)
```

### Interpretation

-   **Country homophily** — same-country artists collaborated significantly more frequently.
-   **Genre homophily** — artists sharing a genre group were more likely to collaborate.
-   **Collaboration count** — highly collaborative artists formed additional ties within the network.
-   **GWESP (triadic closure)** — the network exhibited strong clustering; artists tended to collaborate within shared collaboration circles.

## Model Diagnostics

```{r}
#| label: diagnostics
#| fig-width: 8
#| fig-height: 8

mcmc.diagnostics(model5)
```

## Goodness-of-Fit

```{r}
#| label: gof
#| fig-width: 8
#| fig-height: 8

gof_model5 <- gof(model5)
plot(gof_model5)
```

GOF diagnostics confirm the final model reproduces key structural properties of the observed network, including degree distributions, shared partner structures, and geodesic distances.

## Network Summary

```{r}
#| label: network-summary

network.size(net)
network.edgecount(net)
network.density(net)

deg <- degree(net)
summary(deg)

components <- sna::component.dist(net)
components
```

## Visualizing the Network

Nodes are colored by genre group, with labels shown for the most highly collaborative artists (degree \> 30).

```{r}
#| label: network-plot
#| fig-width: 8
#| fig-height: 8

genre_colors <- c(
  "Hip-Hop"    = "red",
  "Pop"        = "blue",
  "Latin"      = "green",
  "Electronic" = "purple",
  "Country"    = "orange",
  "Other"      = "gray"
)

deg <- sna::degree(net)
node_labels <- ifelse(deg > 30, network.vertex.names(net), "")

plot(net,
     mode        = "fruchtermanreingold",
     vertex.col  = genre_colors[net %v% "genre_group"],
     vertex.cex  = 1.5,
     edge.col    = "#D3D3D3",
     label       = node_labels,
     label.cex   = 0.7)

legend("topright",
       legend = names(genre_colors),
       col    = genre_colors,
       pch    = 19,
       pt.cex = 1.5,
       bty    = "n")
```

## Conclusion

The nested ERGM successfully captured both structural and attribute-based mechanisms underlying artist collaborations. Results suggest that collaboration patterns in the music industry are influenced by shared characteristics such as country and genre, as well as endogenous network processes including clustering and repeated partnership formation.

## Future Work

-   Incorporating temporal dynamics to examine how collaboration patterns evolve over time
-   Expanding the network to include additional artist attributes such as record labels and streaming popularity
-   Modeling weighted collaborations based on collaboration frequency or commercial success of tracks
